PA1_template.Rmd
=============================================================================

Loading the data

```{r Load the data, echo = TRUE}
activity <- read.csv("activity.csv")
```

preprocessing the data

```{r Process/transform the data (if necessary) into a format suitable for your analysis, echo = TRUE}
activity$date <- as.POSIXct(activity$date, "%Y-%m-%d", format = "%Y-%m-%d", tz = "")
weekday <- weekdays(activity$date)
activity <- cbind(activity,weekday)
str(activity)
```

What is mean total number of steps taken per day?

Calculate the total number of steps taken per day

```{r total number of steps taken per day, echo = TRUE}
stepsperday <- with(activity, aggregate(steps, by = list(date), FUN = sum, na.rm = TRUE))
colnames(stepsperday) <- c("date", "steps")
stepsperday
```

Make a histogram of the total number of steps taken each day

```{r histogram with the frequency ot total numbers, echo = TRUE}
hist(stepsperday$steps, main = "histogram of the total number of steps taken each day", xlab = "total number of steps taken each day")
```

Calculate and report the mean of the total number of steps taken per day

```{r mean of the total number of steps, echo = TRUE}
mymean <- mean(stepsperday$steps)
mymean
```

Calculate and report the median of the total number of steps taken per day

```{r median of the total number of steps, echo = TRUE}
mymedian <- median(stepsperday$steps)
mymedian
```

What is the average daily activity pattern?

Make a time series plot of the 5-minute interval and the average number of steps taken, averaged across all days

```{r average daily activity pattern, echo = TRUE}
averagestepsperday <- aggregate(activity$steps, by=list(activity$interval), FUN=mean, na.rm=TRUE)
names(averagestepsperday) <- c("interval", "mean")
plot(averagestepsperday$interval, averagestepsperday$mean, type = "l", xlab="5-minute interval", ylab="average number of steps all days", main="average number of steps per 5-minute interval")

```

Which 5-minute interval, on average across all the days in the dataset, contains the maximum number of steps?

```{r define the interval with the maximum number of steps, echo = TRUE}
averagestepsperday[which.max(averagestepsperday$mean), ]$interval
```

Imputing missing values

Calculate and report the total number of missing values in the dataset

```{r total number of NA values in the dataset, echo = TRUE}
mysumNA <- sum(is.na(activity$steps) == TRUE)
mysumNA
```

Create a new dataset that is equal to the original dataset but with the missing data filled in

```{r new dataset with the missing data filled in, echo = TRUE}
newactivity <- averagestepsperday$mean[match(activity$interval, averagestepsperday$interval)]
newactivity <- transform(activity, steps = ifelse(is.na(activity$steps), yes = newactivity, no = activity$steps))
allactivity <- aggregate(steps ~ date, newactivity, sum)
colnames(allactivity) <- c("date", "stepsbyday")
allactivity
```

Make a histogram of the total number of steps taken each day

```{r histogram with the new frequencies of total number of steps after imputing, echo = TRUE}
hist(allactivity$stepsbyday, xlab = "total number of steps taken each day", main = "histogram of the total number of steps taken each day")
```

Calculate and report the mean of the total number of steps taken per day

```{r new mean of the total number of steps after imputing, echo = TRUE}
mynewmean <- mean(allactivity$stepsbyday)
mynewmean
```

Calculate and report the median of the total number of steps taken per day

```{r new median of the total number of steps after imputing, echo = TRUE}
mynewmedian <- median(allactivity$stepsbyday)
mynewmedian
```

Are there differences in activity patterns between weekdays and weekends?

Create a new factor variable in the dataset with two levels – “weekday” and “weekend” indicating whether a given date is a weekday or weekend day

```{r new variable in the dataset indicating whether a given date is a weekday or weekend day, echo = TRUE}
activity$date <- as.Date(strptime(activity$date, format="%Y-%m-%d"))
activity$datetype <- sapply(activity$date, function(x) {
        if (weekdays(x) == "Saturday" | weekdays(x) =="Sunday") 
                {y <- "Weekend"} else 
                {y <- "Weekday"}
                y
        })
```

Make a panel plot containing a time series plot of the 5-minute interval and the average number of steps taken, averaged across all weekday days or weekend days

```{r plot the number of steps for all 5-min intervals, averaged across weekdays and weekends separately, echo = TRUE}
library(ggplot2)
activitydate <- aggregate(steps~interval + datetype, activity, mean, na.rm = TRUE)
plot<- ggplot(activitydate, aes(x = interval , y = steps, color = datetype)) +
       geom_line() +
       labs(title = "Average daily steps by type of date", x = "Interval", y = "Average number of steps") +
       facet_wrap(~datetype, ncol = 1, nrow=2)
print(plot)
```
